---
title: 'Crack-Seal Veins Data Clean up'
author: 'Daniel Pliego'
date: " March, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
library(readxl)
library(dplyr)
```

The main thing from this file is the get_dataset() function


### Dataset

First I want to make a single data set with all the different samples and separate them by sample

There are a total of 6 samples:

profile 1

profile 2, 3, 4, 5

profile 6, 7

profile 8

profile 9

profile 10

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# I will make a function here that I can call from from somewhere else to get
# The data

get_dataset <- function() {
# First sample
data_cracks_1 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                          skip = 1, sheet = 'Profile 1')

thickness_values_1 <- data_cracks_1$`thickness (um)`

thickness_values_1 <- thickness_values_1[1:(length(thickness_values_1) - 3)]



# Second sample
data_cracks_2 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                            sheet = 'Profiles 2, 3, 4, 5')

thickness_values_2 <- na.omit(data_cracks_2$Thickness)

# Third sample
data_cracks_3_1 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                            sheet = 'Profile 6', skip = 2)

data_cracks_3_2 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                            sheet = 'Profile 7', skip = 2)

thickness_values_3_1 <-  na.omit(data_cracks_3_1$thickness)

thickness_values_3_2 <-  na.omit(data_cracks_3_2$Thickness)


thickness_values_3 <- c(thickness_values_3_1, thickness_values_3_2)




# Fourth sample
data_cracks_4 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                            sheet = 'Profile 8', skip = 2)

thickness_values_4 <- na.omit(data_cracks_4$Thickness)

# Fifth sample
data_cracks_5 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                            sheet = 'Profile 9', skip = 2)

thickness_values_5 <- na.omit(data_cracks_5$Thickness)


# Sixth sample
data_cracks_6 <- read_excel("/Users/pliego/Downloads/HC05_crack_seal_statistics_optical ms.xlsx",
                            sheet = 'Profile 10', skip = 2)

thickness_values_6 <- na.omit(data_cracks_6$Thickness)


## Now we merge everything into a single dataset and label them by sample

thickness_data <- data.frame(
  Thickness = c(thickness_values_1, thickness_values_2, thickness_values_3,
                thickness_values_4, thickness_values_5, thickness_values_6),
  Sample = c(rep(1, length(thickness_values_1)),
             rep(2, length(thickness_values_2)),
             rep(3, length(thickness_values_3)),
             rep(4, length(thickness_values_4)),
             rep(5, length(thickness_values_5)),
             rep(6, length(thickness_values_6)))
)
return(thickness_data)
}

```





#####################################################################################
This is unrelated to the project

### Time series for each sample

```{r, echo=TRUE, fig.height=12, fig.width=8}
#Adjust plots
par(mfrow=c(3,2), mar=c(4,4,2,1)) 

thickness_data <- get_dataset()

# Create a list of time series for each sample
samples <- 1:6
ts_list <- lapply(samples,
                  function(i) ts(thickness_data$Thickness[thickness_data$Sample == i]))

# Plot each time series
for (i in seq_along(ts_list)) {
  plot(ts_list[[i]], main=paste("Sample", samples[i], "Plot"),
       ylab="Thickness", xlab="Index",lwd=2)
}

```


Density plots foe each sample


```{r}


par(mfrow=c(3,2), mar=c(4,4,2,1))

thickness_data <- get_dataset()

# Create a list of thickness values for each sample
samples <- 1:6
ts_list <- lapply(samples, function(i) thickness_data$Thickness[thickness_data$Sample == i])

# Plot density for each sample
for (i in seq_along(ts_list)) {
  dens <- density(ts_list[[i]], bw=2)  # bw=2 smooths over every 2 integer values
  plot(dens, main=paste("Sample", samples[i], "Density"), 
       xlab="Thickness", ylab="Density", lwd=2)
}
```





### Correlograms for each sample

```{r, echo=TRUE, fig.height=12, fig.width=8}
#Adjust plots
par(mfrow=c(3,2), mar=c(4,4,2,1)) 



# Generate correlograms for each sample
for (i in seq_along(ts_list)) {
  acf(ts_list[[i]], main=paste("Sample", samples[i], "Correlogram"), lwd=2)
}


```






Playing around

```{r, echo=TRUE}

spec1 <- spec.pgram(ts_list[[2]], main = "Raw Periodogram of WW", log = "no")


 spec1 <- spec.pgram(ts_list[[2]], main = "Raw Periodogram of WW", log = "no", spans = c(25))
 spec1 <- spec.pgram(ts_list[[3]], main = "Raw Periodogram of WW", log = "no", c(15))

max_indexes <- numeric(0)
max_freqs <- numeric(0)

max_index <- which.max(spec1$spec)

max_freq <- spec1$freq[max_index]


#spec_2 <- spec.pgram(ts_list[[2]], log = "no", spans = c(8,10))

for (i in seq_along(ts_list)) {
  periodogram <- spec.pgram(ts_list[[i]], main = "Raw Periodogram of WW", log = "no",
             spans = c(9))
  
  max_indexes[i] <-  which.max(periodogram$spec)
  max_freqs[i] <- periodogram$freq[max_index]
  
  periodogram
}

```

```{r, echo=TRUE}
#Adjust plots
N <- length(ts_list[[2]])
# data frame to store results
results <- data.frame(p = numeric(0),
                      omega = numeric(0),
                      F_statistic = numeric(0),
                      p_value = numeric(0))


# For loop to fit N/2 models and find the significant p's
for (p in 1:(N/2)) {
  t <- 1:N 
  omega_p <- 2 * pi * p / N 
  
  # Fit the model
  linear_model <- lm(ts_list[[2]] ~ cos(omega_p * t) + sin(omega_p * t))
  
  
  # Get the F-statistic from the model summary
  model_summary <- summary(linear_model)
  F_stat <- model_summary$fstatistic[1]
   # Degrees of freedom for the model 
  df1 <- model_summary$fstatistic[2]
   # Residual degrees of freedom 
  df2 <- model_summary$fstatistic[3] 
  
  # get the p-value 
  p_value <- pf(F_stat, df1, df2, lower.tail = FALSE)
  
  # 99% significance level  
if (p_value < 0.05) {
    results <- rbind(results,
                     data.frame(p = p,
                                omega = omega_p,
                                F_statistic = F_stat,
                                p_value = p_value))
  }
} 
results
```







